Site-Heterogeneous Character Change Models for Morphology

### 2018-12-16 Papers/DraftOutline_SSCCM.tex
adding to discussion

	Unlike a matrix of nucleotide characters, a `0' character at one state may mean that the trait never evolved in one lineage, but was lost in another.
	This lack of common meaning complicates the interpretation of how change occurs over time.
	Likewise, the magnitude of change from one state to another may be different between characters, even if the characters mean the same thing (i.e. `0' representing loss and `1' representing gain).
	For example, in one character, changing from a `0' state to a `1' state could represent a change at a single locus, but at another could represent a much larger change coordinated across many genes.

@@

\subsubsection{Model comparison}

The appropriate number of categories $k$ is selected by comparing estimates of the marginal likelihood under different values for $k$.
Marginal likelihoods were estimated using the steppingstone sampling method \citep{xie2010improving} implemented in RevBayes.

\subsection{Data Matrices}
\subsubsection{Empirical Matrices}
Several large and well-documented ant matrices were used in this study.
The first was that of Keller (2011). 
This matrix is of extant ant groups. 
Keller's matrix was collected with special attention to the poneromorph subgroups (Amblyoponinae, Ectatomminae,
Heteroponerinae, Paraponerinae, Ponerinae, and Proceratiinae) and has 139 characters and 105 taxa.
The matrix is highly complete for a morphological dataset, with about 60\% of cells being filled.
Of the total character set, 100 characters were binary and 39 characters were multistate.
Because of the scope of Keller's study, the taxon sampling is biased towards the poneromorphs, and away from the other ant subfamilies, including large subfamilies such as the Dolichoderinae and the Formicinae. \par
We also used data from Barden and Grimaldi (2015). 
This matrix contained both extant and extinct ants, and was collected in order to place stem ants from the Cretaceous period.
One crown ant amber fossil, \textit{Kyromyrma neffi}, was included in this matrix, but the key feature of this matrix is the sampling of the stem group, including multiple samples from the \textit{Gerontoformica} genus.
This matrix also expands sampling in the non-ponerine groups that are underrepresented in the Keller matrix.
The Barden-Grimaldi matrix contained 42 characters and 41 taxa. 
Of these characters, 26 were binary.\par
In order to achieve maximum coverage, we merged these character matrices.
The taxonomic overlap between the two was 11 taxa, and all subfamilies of the Formicidae except the extinct Formiciinae were represented in the matrix. 
Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
Six characters were recoded to make the character states uniform between the two matrices.
Most of these changes simply involved changing the terminology used in reference to the character states.
For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. 
We will refer to this dataset as the combined dataset. \par
We tested the SHDM model on both binary and binary + multistate datasets.
First, to test the model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
In both cases, we performed Bayes Factor model fitting to determine how many discrete Beta categories best model the data. \par
To test the full SHDM model, we used both the Barden-Grimaldi dataset and the combined dataset.
In testing the SHDM model, we modeled the binary and multistate characters together.
This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data to the phylogenetic question. 
\par

@@

Because the true tree cannot be known from empirical data, we used simulated data to observe the effect this model has on accuracy of estimation, and to see if it is possible to detect the true number of discrete categories of state frequency variation.
To this end, we estimated trees for each replicate under the true model.
We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). \par

@@

\subsection{Simulated Phylogenetic Analyses}
\subsection{Binary 4-cat small}

\subsection{MS4cat small}

\subsection{MS4cat large}

@@

	a) How has the modeling of morphology changed since the Mk was proposed by Lewis? What aspects are comparable to how we model DNA? Which are not? What are potentially interesting future directions? 
	b) General discussion of usage of treeset visualization
	c) something else.

\section{Conclusion}

>>>>

	Unlike a matrix of nucleotide characters, in which a state carries common meaning across sites, a `1' state at one character may represent a derived state requiring many underlying genetic changes.
	At another character, it could represent a derived state requiring only one underlying genetic change.
	Characters can also differ in meanings with respect to whether they are coded as ancestral or derived.
	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher.
	This may not be the case for composite matrices or matrices coded without specific reference to an outgroup or character polarities.
	This lack of common meaning limits a researcher's ability to specify a common mechanism.

@@

\subsection{Data Matrices}
\subsubsection{Empirical Matrices}
Several large and well-documented ant matrices were used in this study.
The first was that of Keller (2011). 
This matrix is of extant ant groups. 
Keller's matrix was collected with special attention to the poneromorph subgroups (Amblyoponinae, Ectatomminae,
Heteroponerinae, Paraponerinae, Ponerinae, and Proceratiinae) and has 139 characters and 105 taxa.
The matrix is highly complete for a morphological dataset, with about 60\% of cells being filled.
Of the total character set, 100 characters were binary and 39 characters were multistate.
Because of the scope of Keller's study, the taxon sampling is biased towards the poneromorphs, and away from the other ant subfamilies, including large subfamilies such as the Dolichoderinae and the Formicinae. \par
We also used data from Barden and Grimaldi (2015). 
This matrix contained both extant and extinct ants, and was collected in order to place stem ants from the Cretaceous period.
One crown ant amber fossil, \textit{Kyromyrma neffi}, was included in this matrix, but the key feature of this matrix is the sampling of the stem group, including multiple samples from the \textit{Gerontoformica} genus.
This matrix also expands sampling in the non-ponerine groups that are underrepresented in the Keller matrix.
The Barden-Grimaldi matrix contained 42 characters and 41 taxa. 
Of these characters, 26 were binary.\par
In order to achieve maximum coverage, we merged these character matrices.
The taxonomic overlap between the two was 11 taxa, and all subfamilies of the Formicidae except the extinct Formiciinae were represented in the matrix. 
Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
Six characters were recoded to make the character states uniform between the two matrices.
Most of these changes simply involved changing the terminology used in reference to the character states.
For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. 
We will refer to this dataset as the combined dataset. \par
We tested the SHDM model on both binary and binary + multistate datasets.
First, to test the model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
In both cases, we performed Bayes Factor model fitting to determine how many discrete Beta categories best model the data. \par
To test the full SHDM model, we used both the Barden-Grimaldi dataset and the combined dataset.
In testing the SHDM model, we modeled the binary and multistate characters together.
This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data to the phylogenetic question. 
\par
\subsubsection{Model comparison}

The appropriate number of categories $k$ for each of the two empirical datasets was selected by comparing estimates of the marginal likelihood under different values for $k$.
Marginal likelihoods were estimated using the steppingstone sampling method \citep{xie2010improving} implemented in RevBayes.

@@

Because the true tree cannot be known from empirical data, we used simulated data to observe the effect this model has on accuracy of estimation, and to see if it is possible to detect the true number of discrete categories of state frequency variation.
To this end, we estimated trees for each replicate under the true model.
We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). 
Estimation was performed in RevBayes under the generating model.\par

@@

\subsection{Simulated Phylogenetic Analyses}
\subsubsection{Binary Data}

In the large datasets, underparameterized models have a deleterious impact on the accuracy of phylogenetic analyses. 
However, overparameterized models did not show a similar decrease in accuracy.
In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. \par

\subsubsection{Multistate Data}

In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparamterization did not. \par

@@

The discussion surrounding the incorporation of morphological data into phylogenetic analyses has largely focused on two endpoints on a spectrum: Parsimony and the Mk model \citep{lewis2001likelihood}.
The idea behind parsimony is simple: that each character can have its own tree and length (number of steps) on that tree.
The simplicity of this idea belies the complexity of the implications about character evolution.
Parsimony can be written out as a likelihood model, referred to as the No Common Mechanisms model.
However, it has been demonstrated that this model is so complex, and features so many parameters	that it is never chosen by even a liberal information criterion. \par
In many ways, the Mk model is similar to an unweighted parsimony model, assuming that the transitions between any two states are equally likely.
But the Mk model assumes a common mechanism across sites, a generalized Jukes-Cantor model.
Application of among site rate variation (ASRV) across characters allows characters to have differing rates of evolution, accommodating natural variation in phenotypic characters. 
Allowing the rate of evolution to vary is known to be benefit in phylogenetic analysis \citep{felsenstein78}, and can potentially be more important than an incorrectly-specified model of evolution \citep{lemmon2004importance}.
Previous work in morphological phylogenetics has also highlighted the importance of appropriate assumptions about rate heterogeneity \citep{Harrison2014}.	\par
The model discussed here is an effort to create a similar framework to ASRV, but for the actual model of evolution. 
We may know little about the probability of a particular transition between two character states \textit{a priori}, in the same way that we may know little about the rate of evolution of any particular site. 
Under the discrete Beta model and the SHDM, state frequencies are drawn from either a Beta (binary data) or Dirichlet (multistate data) prior distribution.
The site likelihoods are then computed by marginalizing over the value of the vector of state frequencies, $\pi_i$.
Mechanistically, this is very similar to the way in which ASRV is calculated. \par
In practice, these model allows us to assume multiple possible mechanisms across a single dataset. 
Previous work on similar models in the software MrBayes \citep{Wright2016} has found that the Mk model is supported using steppingstone model selection in about half of phylogenetic datasets tested.
However, for modeling the other half, a single common mechanism is not adequate. 
In this paper, we have implemented the Beta model from MrBayes, and clarified a model for multistate characters to allow modeling of evolution when a single, common mechanism cannot be assumed. \par
Trees estimated from the empirical data under parsimony are extremely incompletely resolved unless clade constraints are used.
This is also the case for our model-based estimations. 
Due to the presence of stem ants the matrices used, particularly those of Barden, would be expected to show multiple models of asymmetric transitions. 
In some characters, we would expect to see a strong signature of loss: stem ants often display characters similar to that of the wasp outgroup, which crown ants do not.
We would expect to see these characters lost, but not regained as expected under the Mk model.
Likewise, due to the presence of ant apomorphies in the crown, we expect to see these characters gained and never lost.
Together, these character classes argue strongly for a symmetric Beta distribution, in which if there is a set of characters with a certain bias (for example, more `0' to `1' transitions), it is expected that there would be another set of characters with an opposite bias. \par
Indeed, each dataset strongly supports the use of a non-Mk model, with differing numbers of categories (Fig. 1).
Why then would a model statistically supported to be a better fit to the data not lead to a more resolved tree?
There are several possible explanations for this.
First, ants are known to have strong convergence within characters.
The group is large and diverse, but fairly morphologically constrained.
While likelihood-based methods are expected to fare better than parsimony when convergence is an issue, they are not perfect and can be mislead, particularly in the presence of model violations such as structured missing data.
Secondly, the dataset is small relative to the number of taxa.
It's possible that there simply isn't enough data to address every split in the tree.
Lastly, I dunno. Morphology what. \par

 \par
